---
title: "Integrated RNA and DNA methylation analysis: effect of acetate and fibre on heart and spleen DNA methylation"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
theme: cosmo
---

Source: https://github.com/markziemann/fran_mbd/blob/master/main_report_integration.Rmd

## Introduction

I will perform a multi-contrast pathway analysis involving RNA expression and DNA methylation data looking
at differences caused by acetate and high fibre diet.

```{r,packages}
suppressPackageStartupMessages({
    # to be run in R4
    library("mitch")
})
```

## Import profiling data 

Importing the DESeq2 tables I generated earlier.

```{r,import}
rna_heart_ctrl_vs_acetate <- read.table("heart_ctrl_vs_acetate_rna.tsv")
rna_heart_ctrl_vs_hifibre <- read.table("heart_ctrl_vs_hifibre_rna.tsv")
rna_spleen_ctrl_vs_acetate <- read.table("spleen_ctrl_vs_acetate_rna.tsv")
rna_spleen_ctrl_vs_hifibre <- read.table("spleen_ctrl_vs_hifibre_rna.tsv")

meth_heart_ctrl_vs_acetate <- read.table("heart_ctrl_vs_acetate_meth.tsv")
meth_heart_ctrl_vs_hifibre <- read.table("heart_ctrl_vs_hifibre_meth.tsv")
meth_spleen_ctrl_vs_acetate <- read.table("spleen_ctrl_vs_acetate_meth.tsv")
meth_spleen_ctrl_vs_hifibre <- read.table("spleen_ctrl_vs_hifibre_meth.tsv")

# strip the gene names out to make downstream analysis easier
rownames(rna_heart_ctrl_vs_acetate) <- sapply(strsplit(rownames(rna_heart_ctrl_vs_acetate),"_"),"[[",1)
rownames(rna_heart_ctrl_vs_hifibre) <- sapply(strsplit(rownames(rna_heart_ctrl_vs_hifibre),"_"),"[[",1)
rownames(rna_spleen_ctrl_vs_acetate) <- sapply(strsplit(rownames(rna_spleen_ctrl_vs_acetate),"_"),"[[",1)
rownames(rna_spleen_ctrl_vs_hifibre) <- sapply(strsplit(rownames(rna_spleen_ctrl_vs_hifibre),"_"),"[[",1)

rownames(meth_heart_ctrl_vs_acetate) <- sapply(strsplit(rownames(meth_heart_ctrl_vs_acetate),"_"),"[[",1)
rownames(meth_heart_ctrl_vs_hifibre) <- sapply(strsplit(rownames(meth_heart_ctrl_vs_hifibre),"_"),"[[",1)
rownames(meth_spleen_ctrl_vs_acetate) <- sapply(strsplit(rownames(meth_spleen_ctrl_vs_acetate),"_"),"[[",1)
rownames(meth_spleen_ctrl_vs_hifibre) <- sapply(strsplit(rownames(meth_spleen_ctrl_vs_hifibre),"_"),"[[",1)
```

## Pathway level analysis with mitch

## First fetch gene sets from Reactome.

```{r,genesets}
# gene sets
download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", 
    destfile="ReactomePathways.gmt.zip")
unzip("ReactomePathways.gmt.zip",overwrite = TRUE)
genesets <- gmt_import("ReactomePathways.gmt")

# gene table
m2h <- read.table("mouse2human.txt.sort")
m2h[,1]=NULL
```

## Heart: ctrl vs acetate

```{r,mitch1}
x <- list("rna"=rna_heart_ctrl_vs_acetate,"meth"=meth_heart_ctrl_vs_acetate)
m <- mitch_import(x=x, DEtype="deseq2",geneTable=m2h)

capture.output(
    res <- mitch_calc(m,genesets=genesets,priority="effect")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

head(res$enrichment_result,20)

unlink("heart_ctrl_vs_acetate_integrate.html")
    capture.output(
        mitch_report(res,outfile=paste("heart_ctrl_vs_acetate_integrate.html"))
        , file = "/dev/null", append = FALSE,
        type = c("output", "message"), split = FALSE)

```

## Heart: ctrl vs high fibre

```{r,mitch2}
x <- list("rna"=rna_heart_ctrl_vs_hifibre,"meth"=meth_heart_ctrl_vs_hifibre)
m <- mitch_import(x=x, DEtype="deseq2",geneTable=m2h)

capture.output(
    res <- mitch_calc(m,genesets=genesets,priority="effect")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

head(res$enrichment_result,20)

unlink("heart_ctrl_vs_hifibre_integrate.html")
    capture.output(
        mitch_report(res,outfile=paste("heart_ctrl_vs_hifibre_integrate.html"))
        , file = "/dev/null", append = FALSE,
        type = c("output", "message"), split = FALSE)

```

## Spleen: ctrl vs acetate 

```{r,mitch3}
x <- list("rna"=rna_spleen_ctrl_vs_acetate,"meth"=meth_spleen_ctrl_vs_acetate)
m <- mitch_import(x=x, DEtype="deseq2",geneTable=m2h)

capture.output(
    res <- mitch_calc(m,genesets=genesets,priority="effect")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

head(res$enrichment_result,20)

unlink("spleen_ctrl_vs_acetate_integrate.html")
    capture.output(
        mitch_report(res,outfile=paste("spleen_ctrl_vs_acetate_integrate.html"))
        , file = "/dev/null", append = FALSE,
        type = c("output", "message"), split = FALSE)

```

## Spleen: ctrl vs high fibre

```{r,mitch4}
x <- list("rna"=rna_spleen_ctrl_vs_hifibre,"meth"=meth_spleen_ctrl_vs_hifibre)
m <- mitch_import(x=x, DEtype="deseq2",geneTable=m2h)

capture.output(
    res <- mitch_calc(m,genesets=genesets,priority="effect")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

head(res$enrichment_result,20)

unlink("spleen_ctrl_vs_hifibre_integrate.html")
    capture.output(
        mitch_report(res,outfile=paste("spleen_ctrl_vs_hifibre_integrate.html"))
        , file = "/dev/null", append = FALSE,
        type = c("output", "message"), split = FALSE)

```

End of report

# Session information

```{r,session}
sessionInfo()
```
