---
title: "MBD-seq analysis: effect of acetate and fibre on heart and spleen DNA methylation"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
theme: cosmo
---


## Introduction

In this report, I will take you through an analysis of the MBD-seq that was previosly generated.
I will perform an analysis that begins with DESeq2 for differential methylation followed by pathway enrichment analysis with mitch.

```{r,packages}
suppressPackageStartupMessages({
    # to be run in R4
    library("tidyverse")
    library("reshape2")
    library("DESeq2")
    library("gplots")
    library("fgsea")
    library("mitch")
})
```

## Import profiling data 

This analysis uses the 3kbp area arount the TSS.

```{r,import}
hm <- read.table("heart_meth_3kbpTSS.txt",header=TRUE,row.names=1)
sm <-read.table("spleen_meth_3kbpTSS.txt",header=TRUE,row.names=1)
b <- data.frame(sm,hm)
```

# MDS plots
```{r,mds}
plot(cmdscale(dist(t(b))), bty='l', xlab="Coordinate 1", ylab="Coordinate 2", 
 type = "p", pch=19, cex.axis=1.3, cex.lab=1.3 , col="gray") 
text(cmdscale(dist(t(b))), labels=colnames(b),cex=1.3) 
mtext("spleen and heart samples")

plot(cmdscale(dist(t(hm))), bty='l', xlab="Coordinate 1", ylab="Coordinate 2",
 type = "p", pch=19, cex.axis=1.3, cex.lab=1.3 , col="gray")
text(cmdscale(dist(t(hm))), labels=colnames(hm),cex=1.3)
mtext("heart samples")

plot(cmdscale(dist(t(sm))), bty='l', xlab="Coordinate 1", ylab="Coordinate 2",
 type = "p", pch=19, cex.axis=1.3, cex.lab=1.3 , col="gray")
text(cmdscale(dist(t(sm))), labels=colnames(sm),cex=1.3)
mtext("spleen samples")
```

# Curate the samplesheet
```{r,samplesheet}
ss_sm <- as.data.frame(colnames(sm))
colnames(ss_sm) <- "samplename"
ss_sm$group  <- sapply(strsplit(as.character(ss_sm$samplename) , "_"),"[[",2)
ss_sm$group <- gsub("[0-9]+","",ss_sm$group)
rownames(ss_sm) <- ss_sm$samplename
ss_sm$samplename=NULL

ss_hm <- as.data.frame(colnames(hm))
colnames(ss_hm) <- "samplename"
ss_hm$group  <- sapply(strsplit(as.character(ss_hm$samplename) , "_"),"[[",2)
ss_hm$group <- gsub("[0-9]+","",ss_hm$group)
rownames(ss_hm) <- ss_hm$samplename
ss_hm$samplename=NULL
```

# Differential analysis

We will be running some comparisons between groups:

1. HEART ctrl vs acetate

2. HEART ctrl vs hifib

3. SPLEEN ctrl vs acetate

4. SPLEEN ctrl vs hifibre


## HEART ctrl vs acetate

```{r,dm1}
x <- hm[,grep("hifi" , colnames(hm) ,invert=TRUE)]
x<-x[which(rowSums(x)/ncol(x)>=(10)),]
xs <- ss_hm[which( rownames(ss_hm) %in% colnames(x) ),1,drop=FALSE]
xs$trt <- factor(as.numeric(grepl("acetate",xs$group)))
dds <- DESeqDataSetFromMatrix(countData = x , colData = xs, design = ~ trt )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE)
zz<-cbind(as.data.frame(z),assay(vsd))
dge<-as.data.frame(zz[order(zz$pvalue),])
write.table(dge,file="heart_ctrl_vs_acetate_deseq.tsv",quote=F,sep="\t")
#some plots
sig<-subset(dge,padj<0.05)
SIG=nrow(sig)
DN=nrow(subset(sig,log2FoldChange<0))
UP=nrow(subset(sig,log2FoldChange>0))
HEADER=paste("ctrl vs acetate:", SIG , "DGEs,", UP ,"up,", DN, "down")
plot(log2(dge$baseMean),dge$log2FoldChange,cex=0.6,cex.axis=1.2,cex.lab=1.3, 
 xlab="log2 base mean",
 ,ylab="log2 fold change" ,pch=19,col="#838383")
points(log2(sig$baseMean),sig$log2FoldChange,cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
top<-head(sig,20)
plot(dge$log2FoldChange, -log2(dge$pvalue)+1E-307 ,cex=0.6, cex.lab=1.3,cex.axis=1.2,
 xlim=c(-3,3),xlab="log2 fold change", ylab="-log2 p-value" ,pch=19,col="#838383")
points(sig$log2FoldChange, -log2(sig$pvalue)+1E-307, cex=0.6,pch=19,col="red")  
mtext((HEADER),cex=1.2)
# top N gene heatmap
colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(  as.matrix(dge[1:50,c(7:ncol(dge))]), col=colfunc(25),scale="row",
 trace="none",margins = c(6,20), cexRow=.6, cexCol=.8,  main="Top 50 genes")

heart_ctrl_vs_acetate <- dge
```

## HEART ctrl vs hifib

```{r,dm2}
x <- hm[,grep("acet" , colnames(hm) ,invert=TRUE)]
x<-x[which(rowSums(x)/ncol(x)>=(10)),]
xs <- ss_hm[which( rownames(ss_hm) %in% colnames(x) ),1,drop=FALSE]
xs$trt <- factor(as.numeric(grepl("hifi",xs$group)))
dds <- DESeqDataSetFromMatrix(countData = x , colData = xs, design = ~ trt )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE)
zz<-cbind(as.data.frame(z),assay(vsd))
dge<-as.data.frame(zz[order(zz$pvalue),])
write.table(dge,file="heart_ctrl_vs_hifibre_deseq.tsv",quote=F,sep="\t")
#some plots
sig<-subset(dge,padj<0.05)
SIG=nrow(sig)
DN=nrow(subset(sig,log2FoldChange<0))
UP=nrow(subset(sig,log2FoldChange>0))
HEADER=paste("ctrl vs hifibre:", SIG , "DGEs,", UP ,"up,", DN, "down")
plot(log2(dge$baseMean),dge$log2FoldChange,cex=0.6,cex.axis=1.2,cex.lab=1.3,
 xlab="log2 base mean",
 ,ylab="log2 fold change" ,pch=19,col="#838383")
points(log2(sig$baseMean),sig$log2FoldChange,cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
top<-head(sig,20)
plot(dge$log2FoldChange, -log2(dge$pvalue)+1E-307 ,cex=0.6, cex.lab=1.3,cex.axis=1.2,
 xlim=c(-3,3),xlab="log2 fold change", ylab="-log2 p-value" ,pch=19,col="#838383")
points(sig$log2FoldChange, -log2(sig$pvalue)+1E-307, cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
# top N gene heatmap
colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(  as.matrix(dge[1:50,c(7:ncol(dge))]), col=colfunc(25),scale="row",
 trace="none",margins = c(6,20), cexRow=.6, cexCol=.8,  main="Top 50 genes")

heart_ctrl_vs_hifibre <- dge
```

## SPLEEN ctrl vs acetate

```{r,dm3}
x <- sm[,grep("hifi" , colnames(sm) ,invert=TRUE)]
x<-x[which(rowSums(x)/ncol(x)>=(10)),]
xs <- ss_sm[which( rownames(ss_sm) %in% colnames(x) ),1,drop=FALSE]
xs$trt <- factor(as.numeric(grepl("acetate",xs$group)))
dds <- DESeqDataSetFromMatrix(countData = x , colData = xs, design = ~ trt )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE)
zz<-cbind(as.data.frame(z),assay(vsd))
dge<-as.data.frame(zz[order(zz$pvalue),])
write.table(dge,file="spleen_ctrl_vs_acetate_deseq.tsv",quote=F,sep="\t")
#some plots
sig<-subset(dge,padj<0.05)
SIG=nrow(sig)
DN=nrow(subset(sig,log2FoldChange<0))
UP=nrow(subset(sig,log2FoldChange>0))
HEADER=paste("ctrl vs acetate:", SIG , "DGEs,", UP ,"up,", DN, "down")
plot(log2(dge$baseMean),dge$log2FoldChange,cex=0.6,cex.axis=1.2,cex.lab=1.3,
 xlab="log2 base mean",
 ,ylab="log2 fold change" ,pch=19,col="#838383")
points(log2(sig$baseMean),sig$log2FoldChange,cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
top<-head(sig,20)
plot(dge$log2FoldChange, -log2(dge$pvalue)+1E-307 ,cex=0.6, cex.lab=1.3,cex.axis=1.2,
 xlim=c(-3,3),xlab="log2 fold change", ylab="-log2 p-value" ,pch=19,col="#838383")
points(sig$log2FoldChange, -log2(sig$pvalue)+1E-307, cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
# top N gene heatmap
colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(  as.matrix(dge[1:50,c(7:ncol(dge))]), col=colfunc(25),scale="row",
 trace="none",margins = c(6,20), cexRow=.6, cexCol=.8,  main="Top 50 genes")

spleen_ctrl_vs_acetate <- dge

```

## SPLEEN ctrl vs hifibre

```{r,dm4}
x <- sm[,grep("acet" , colnames(sm) ,invert=TRUE)]
x<-x[which(rowSums(x)/ncol(x)>=(10)),]
xs <- ss_sm[which( rownames(ss_sm) %in% colnames(x) ),1,drop=FALSE]
xs$trt <- factor(as.numeric(grepl("hifi",xs$group)))
dds <- DESeqDataSetFromMatrix(countData = x , colData = xs, design = ~ trt )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE)
zz<-cbind(as.data.frame(z),assay(vsd))
dge<-as.data.frame(zz[order(zz$pvalue),])
write.table(dge,file="spleen_ctrl_vs_hifibre_deseq.tsv",quote=F,sep="\t")
#some plots
sig<-subset(dge,padj<0.05)
SIG=nrow(sig)
DN=nrow(subset(sig,log2FoldChange<0))
UP=nrow(subset(sig,log2FoldChange>0))
HEADER=paste("ctrl vs hifibre:", SIG , "DGEs,", UP ,"up,", DN, "down")
plot(log2(dge$baseMean),dge$log2FoldChange,cex=0.6,cex.axis=1.2,cex.lab=1.3,
 xlab="log2 base mean",
 ,ylab="log2 fold change" ,pch=19,col="#838383")
points(log2(sig$baseMean),sig$log2FoldChange,cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
top<-head(sig,20)
plot(dge$log2FoldChange, -log2(dge$pvalue)+1E-307 ,cex=0.6, cex.lab=1.3,cex.axis=1.2,
 xlim=c(-3,3),xlab="log2 fold change", ylab="-log2 p-value" ,pch=19,col="#838383")
points(sig$log2FoldChange, -log2(sig$pvalue)+1E-307, cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
# top N gene heatmap
colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(  as.matrix(dge[1:50,c(7:ncol(dge))]), col=colfunc(25),scale="row",
 trace="none",margins = c(6,20), cexRow=.6, cexCol=.8,  main="Top 50 genes")

spleen_ctrl_vs_hifibre <- dge
```

# Pathway level analysis with mitch

## First fetch gene sets from Reactome.

```{r,genesets}
library("mitch")
# gene sets
download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", 
    destfile="ReactomePathways.gmt.zip")
unzip("ReactomePathways.gmt.zip",overwrite = TRUE)
genesets <- gmt_import("ReactomePathways.gmt")
```

Now I will run enrichment analysis with mitch.
Firstly each contrast individually but also all at once.

```{r,gt}

m2h <- read.table("mouse2human.txt.sort")
m2h[,1]=NULL
```

## Heart: ctrl vs acetate

```{r,mitch1}
rownames(heart_ctrl_vs_acetate) <- sapply(strsplit(rownames(heart_ctrl_vs_acetate),"_"),"[[",1)
m <- mitch_import(x=heart_ctrl_vs_acetate, DEtype="deseq2",geneTable=m2h)

capture.output(
    res <- mitch_calc(m,genesets=genesets,priority="effect")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

head(res$enrichment_result,20)

unlink("heart_ctrl_vs_acetate.html")
    capture.output(
        mitch_report(res,outfile=paste("heart_ctrl_vs_acetate.html"))
        , file = "/dev/null", append = FALSE,
        type = c("output", "message"), split = FALSE)

```

## Heart: ctrl vs high fibre

```{r,mitch2}
rownames(heart_ctrl_vs_hifibre) <- sapply(strsplit(rownames(heart_ctrl_vs_hifibre),"_"),"[[",1)
m <- mitch_import(x=heart_ctrl_vs_hifibre, DEtype="deseq2",geneTable=m2h)

capture.output(
    res <- mitch_calc(m,genesets=genesets,priority="effect")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

head(res$enrichment_result,20)

unlink("heart_ctrl_vs_hifibre.html")
    capture.output(
        mitch_report(res,outfile=paste("heart_ctrl_vs_hifibre.html"))
        , file = "/dev/null", append = FALSE,
        type = c("output", "message"), split = FALSE)

```

## Spleen: ctrl vs acetate 

```{r,mitch3}
rownames(spleen_ctrl_vs_acetate) <- sapply(strsplit(rownames(spleen_ctrl_vs_acetate),"_"),"[[",1)
m <- mitch_import(x=spleen_ctrl_vs_acetate, DEtype="deseq2",geneTable=m2h)

capture.output(
    res <- mitch_calc(m,genesets=genesets,priority="effect")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

head(res$enrichment_result,20)

unlink("spleen_ctrl_vs_acetate.html")
    capture.output(
        mitch_report(res,outfile=paste("spleen_ctrl_vs_acetate.html"))
        , file = "/dev/null", append = FALSE,
        type = c("output", "message"), split = FALSE)

```

## Spleen: ctrl vs high fibre

```{r,mitch4}
rownames(spleen_ctrl_vs_hifibre) <- sapply(strsplit(rownames(spleen_ctrl_vs_hifibre),"_"),"[[",1)
m <- mitch_import(x=spleen_ctrl_vs_hifibre, DEtype="deseq2",geneTable=m2h)

capture.output(
    res <- mitch_calc(m,genesets=genesets,priority="effect")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

head(res$enrichment_result,20)

unlink("spleen_ctrl_vs_hifibre.html")
    capture.output(
        mitch_report(res,outfile=paste("spleen_ctrl_vs_hifibre.html"))
        , file = "/dev/null", append = FALSE,
        type = c("output", "message"), split = FALSE)

```

## Multi enrichment

```{r,multi}
x <- list("heart_ctrl_vs_acetate"=heart_ctrl_vs_acetate,
    "heart_ctrl_vs_hifibre"=heart_ctrl_vs_hifibre,
    "spleen_ctrl_vs_acetate"=spleen_ctrl_vs_acetate,
    "spleen_ctrl_vs_hifibre"=spleen_ctrl_vs_hifibre)

m <- mitch_import(x=x, DEtype="deseq2",geneTable=m2h)

capture.output(
    res <- mitch_calc(m,genesets=genesets,priority="effect")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

head(res$enrichment_result,20)

unlink("multi_enrichment.html")
    capture.output(
        mitch_report(res,outfile=paste("multi_enrichment.html"))
        , file = "/dev/null", append = FALSE,
        type = c("output", "message"), split = FALSE)
```

End of report

# Session information

```{r,session}
sessionInfo()
```
